---
title: "R Notebook"
output: html_notebook
---

# Load Libraries

```{r}
library(dplyr)
library(DHARMa)
library(ggplot2)
library(emmeans)
```

# Load and Merge Data

Load fISH statistics files, making sure to fix incompatible sample ids.
```{r}
dat.raw <- read.csv("../Preprocessed\ Statistics\ Files/nbm_and_vp_stats_merged.csv")
# 4607 -> 4670
dat.raw$sample_id[dat.raw$sample_id == "4607"] <- "4670"
# 5401-1 -> 5401
dat.raw$sample_id[dat.raw$sample_id == "5401-1"] <- "5401"
dat.raw
```

Load specimen info, convert categorical variables to factors, align terminology to better match manuscript, and change name of column for merge.
```{r}
dat.info <- read.csv("../Preprocessed\ Statistics\ Files/specimen\ info_ASD\ genetics.csv")
dat.info$Disorder[dat.info$Disorder == "ASD - Autism"] <- "ASD"
dat.info$Disorder[dat.info$Disorder == "Control"] <- "AST"

colnames(dat.info)[2] = "Neurotype"
colnames(dat.info)[1] = "sample_id"

dat.info$Neurotype <- as.factor(dat.info$Neurotype)
dat.info$Race <- as.factor(dat.info$Race)
dat.info$Sex <- as.factor(dat.info$Sex)

dat.info
```

Left join fISH statistics with specimen info.
```{r}
dat.merged <- merge(x = dat.raw, y = dat.info, by = "sample_id", all.x = T)
dat.merged
```

## Commentary
The columns above correspond to the metrics described in the manuscript.
See the table below for a mapping between variable names and descriptions in the manuscript.

  Full Name (Manuscript)                  Abbreviation (Manuscript)   Column Name (Table)       
---------------------------------------- --------------------------- ---------------------------
  NBM *OXTR* mRNA Brightness Ratio        NB                          `nbm_brightness`          
  NBM *ChAT+ OXTR* mRNA Brightness Ratio  NBRC+                       `nbm_brightness_ratio`
  NBM *ChAT- OXTR* mRNA Brightness Ratio  NBRC-                       `nbm_inv_brightness_ratio`
  VP *OXTR* mRNA Brightness Ratio         VB                          `vp_brightness`

# NB Analysis
We'll now proceeed to analyze our NBM *OXTR* mRNA Brightness metric.

## Exploratory Analysis

First, let's find our sample sizes (split up by group).
```{r}
with(filter(dat.merged, !is.na(nbm_brightness)),
     table(Neurotype, useNA = "ifany"))
```

Now let's plot the data.
First, let's try it without any transforms.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = nbm_brightness)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```

Since the data appear skewed, let's try a log-transform.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = log(nbm_brightness))) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```
Although it's not perfect, NB does appear log-normally distributed.
Based on just visual examination, there does not appear to be a significant difference between ASD and AST for our NB metric, although NB does appear to trend higher for our ASD group.

Finally, let's compute median values by Neurotype.
```{r}
med.nb.orig <- dat.merged %>%
  group_by(Neurotype) %>%
  summarize(median = median(nbm_brightness, na.rm = T)) %>%
  ungroup()
med.nb.orig
```

## Model Fits
Since we have skewed data, we'll try fitting a few different models and evaluate the residuals to choose our best option.

### Non-Transformed
```{r}
mod.nt.nb.n <- lm(nbm_brightness ~ Neurotype, data = dat.merged)
summary(mod.nt.nb.n)
```
Now let's examine the model residuals.
```{r}
mod.nt.nb.n.res <- simulateResiduals(mod.nt.nb.n, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nb.n, ~Neurotype)
```

### Log-Transformed
```{r}
mod.nt.nb.l <- lm(log(nbm_brightness) ~ Neurotype, data = dat.merged)
summary(mod.nt.nb.l)
```

Now let's examine the model residuals.
```{r}
mod.nt.nb.l.res <- simulateResiduals(mod.nt.nb.l, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nb.l, ~Neurotype, type = "response")
```

### Square-Root-Transformed
```{r}
mod.nt.nb.s <- lm(sqrt(nbm_brightness) ~ Neurotype, data = dat.merged)
summary(mod.nt.nb.s)
```

Now let's examine the model residuals.
```{r}
mod.nt.nb.s.res <- simulateResiduals(mod.nt.nb.s, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nb.s, ~Neurotype, type = "response")
```

### Model Selection
Based on the statistics above, we'll be selecting the log-transformed model to evaluate our NB metric. 
This is because it's residuals are approximately normal, while keeping the estimated marginal means close to the median values found above.

---

# NBRC+ Analysis
We'll now proceeed to analyze our NBM *ChAT+ OXTR* mRNA Brightness Ratio metric.

## Exploratory Analysis

First, let's find our sample sizes (split up by group).
```{r}
with(filter(dat.merged, !is.na(nbm_brightness_ratio)),
     table(Neurotype, useNA = "ifany"))
```

Now let's plot the data.
First, let's try it without any transforms.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = nbm_brightness_ratio)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```

Although it's not perfect, NBRC+ does appear normally distributed.
Based on just visual examination, there does not appear to be a significant difference between ASD and AST for our NBRC+ metric, although NBRC+ does appear to trend higher for our ASD group.

Now, let's compute median values by Neurotype.
```{r}
med.nbrcp.orig <- dat.merged %>%
  group_by(Neurotype) %>%
  summarize(median = median(nbm_brightness_ratio, na.rm = T)) %>%
  ungroup()
med.nbrcp.orig
```

## Model Fits
This metric appears to be normally distributed, so let's start with a non-transformed model.

### Non-Transformed
```{r}
mod.nt.nbrcp.n <- lm(nbm_brightness_ratio ~ Neurotype, data = dat.merged)
summary(mod.nt.nbrcp.n)
```
Now let's examine the model residuals.
```{r}
mod.nt.nbrcp.n.res <- simulateResiduals(mod.nt.nbrcp.n, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nbrcp.n, ~Neurotype)
```

### Model Selection
The non-transformed model above looks really good, and it's the one we'll be using to analyze our NBRC+ metric.

---


# NBRC- Analysis
We'll now proceeed to analyze our NBM *ChaT- OXTR* mRNA Brightness Ratio metric.

## Exploratory Analysis

First, let's find our sample sizes (split up by group).
```{r}
with(filter(dat.merged, !is.na(nbm_inv_brightness_ratio)),
     table(Neurotype, useNA = "ifany"))
```

Now let's plot the data.
First, let's try it without any transforms.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = nbm_inv_brightness_ratio)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```

Since the data appear skewed, let's try a log-transform.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = log(nbm_inv_brightness_ratio))) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```
Although it's not perfect, NBRC- does appear log-normally distributed.
Based on just visual examination, there does not appear to be a significant difference between ASD and AST for our NBRC- metric, although NBRC- does appear to trend higher for our ASD group.

Finally, let's compute median values by Neurotype.
```{r}
med.nbrcm.orig <- dat.merged %>%
  group_by(Neurotype) %>%
  summarize(median = median(nbm_inv_brightness_ratio, na.rm = T)) %>%
  ungroup()
med.nbrcm.orig
```

## Model Fits
Since we have skewed data, we'll try fiting a few different models and evaluate the residuals to choose our best option.

### Non-Transformed
```{r}
mod.nt.nbrcm.n <- lm(nbm_inv_brightness_ratio ~ Neurotype, data = dat.merged)
summary(mod.nt.nbrcm.n)
```
Now let's examine the model residuals.
```{r}
mod.nt.nbrcm.n.res <- simulateResiduals(mod.nt.nbrcm.n, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nbrcm.n, ~Neurotype)
```

### Log-Transformed
```{r}
mod.nt.nbrcm.l <- lm(log(nbm_inv_brightness_ratio) ~ Neurotype, data = dat.merged)
summary(mod.nt.nbrcm.l)
```

Now let's examine the model residuals.
```{r}
mod.nt.nbrcm.l.res <- simulateResiduals(mod.nt.nbrcm.l, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nbrcm.l, ~Neurotype, type = "response")
```

### Square-Root-Transformed
```{r}
mod.nt.nbrcm.s <- lm(sqrt(nbm_inv_brightness_ratio) ~ Neurotype, data = dat.merged)
summary(mod.nt.nbrcm.s)
```

Now let's examine the model residuals.
```{r}
mod.nt.nbrcm.s.res <- simulateResiduals(mod.nt.nbrcm.s, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.nbrcm.s, ~Neurotype, type = "response")
```

### Model Selection
Based on the statistics above, we'll be selecting the log-transformed model to evaluate our NBRC- metric. 
This is because it's residuals are approximately normal, while keeping the estimated marginal means close to the median values found above.

---

# VB Analysis
We'll now proceed to analyze our VP *OXTR* mRNA Brightness metric.

## Exploratory Analysis

First, let's find our sample sizes (split up by group).
```{r}
with(filter(dat.merged, !is.na(vp_brightness)),
     table(Neurotype, useNA = "ifany"))
```

Now let's plot the data.
First, let's try it without any transforms.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = vp_brightness)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```

Since the data appear skewed, let's try a log-transform.
```{r}
ggplot(dat.merged,
       aes(x = Neurotype, y = log(vp_brightness))) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.1)
```
Although it's not perfect, VB does appear log-normally distributed.
Based on just visual examination, it appears there may be a significant difference between neurotypes for this metric.

Finally, let's compute median values by Neurotype.
```{r}
med.vb.orig <- dat.merged %>%
  group_by(Neurotype) %>%
  summarize(median = median(vp_brightness, na.rm = T)) %>%
  ungroup()
med.vb.orig
```

## Model Fits
Since we have skewed data, we'll try fiting a few different models and evaluate the residuals to choose our best option.

### Non-Transformed
```{r}
mod.nt.vb.n <- lm(vp_brightness ~ Neurotype, data = dat.merged)
summary(mod.nt.vb.n)
```
Now let's examine the model residuals.
```{r}
mod.nt.vb.n.res <- simulateResiduals(mod.nt.vb.n, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.vb.n, ~Neurotype)
```

### Log-Transformed
```{r}
mod.nt.vb.l <- lm(log(vp_brightness) ~ Neurotype, data = dat.merged)
summary(mod.nt.vb.l)
```

Now let's examine the model residuals.
```{r}
mod.nt.vb.l.res <- simulateResiduals(mod.nt.vb.l, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.vb.l, ~Neurotype, type = "response")
```

### Square-Root-Transformed
```{r}
mod.nt.vb.s <- lm(sqrt(vp_brightness) ~ Neurotype, data = dat.merged)
summary(mod.nt.vb.s)
```

Now let's examine the model residuals.
```{r}
mod.nt.vb.s.res <- simulateResiduals(mod.nt.vb.s, plot = T)
```

And estimate marginal means:
```{r}
emmeans(mod.nt.vb.s, ~Neurotype, type = "response")
```

### Model Selection
Based on the statistics above, we'll be selecting the log-transformed model to evaluate our NBRC- metric. 
This is because it's residuals are approximately normal, while keeping the estimated marginal means close to the median values found above.